---
title: "PREHEAT Data Analysis Report - Machine Learning"
author: "Elise Hickman(primary author), Hannah Matthews, Morgan Nalesnik"
date: "2022-10-21"
output: 
 html_document:
    toc: true
    toc_float: true
    toc_depth: 5
    code_folding: show
---

## PREHEAT Retreat Script Description

Authored by Elise Hickman, Hannah Matthews, and Morgan Nalesnik. Group 5 - Asthma Status.
Aim of script: determine whether machine learning models can accurately determine asthma status using proteins as predictors.

## Load packages and data

```{r message = FALSE}
# Load packages
library(tidyverse)
library(janitor)
library(xgboost)
```
```{r}
# Load data
data_proteomics <- read.csv("./3. Data Analysis/Data/Processed_Proteomics_Data.csv")
data_demo <- read.csv("./3. Data Analysis/Data/Demographic_Data.csv")
```

## Reformat data

Convert to tidy format and create full data frame.

```{r}
# Change to tidy format 
data_tidy <-
  data_proteomics %>% 
  pivot_longer(cols=c(2:ncol(data_proteomics)), names_to = "SampleID", values_to = 'Value') %>% 
  pivot_wider(names_from = 'Gene_Name', values_from='Value')

data_demo <-
  data_demo %>% 
  select(SampleID, Asthma_Status)

data_predict <-
  data_tidy %>% 
  left_join(data_demo, by='SampleID') %>% 
  select(c(2:ncol(data_tidy), Asthma_Status)) %>%
  clean_names() %>% 
  mutate(asthma_status = ifelse(asthma_status == 'NAS', 0, 1))
```

Shuffle data set and split into training and testing datasets.
```{r}
set.seed(6124)
shuffle_idx = sample(nrow(data_predict), nrow(data_predict))
data_predict_shuffled <- data_predict[shuffle_idx,]

split_idx <- sample(c(T,F), replace = T, size=nrow(data_predict), prob=c(0.80, 0.20)) 

data_train <- data_predict_shuffled[split_idx,]
data_test <- data_predict_shuffled[!split_idx,]
```

## Logistic Regression

First, model is trained, then model is tested. Performance parameters can be calculated from prediction results.

```{r warning = FALSE}
# Train model
mdl <- glm(asthma_status~., family=binomial, data=data_train)

# Predict
logit_pred <- predict(mdl, data_test %>% select(-asthma_status), type='response')

# Convert predictions into integers 
logit_bin_pred <- as.integer(logit_pred > 0.5)

# Calculate accuracy
lr_acc <- mean(logit_bin_pred == data_test$asthma_status)
lr_acc

# Calculate specificity
tp <- sum(logit_bin_pred == 1 & data_test$asthma_status == 1)
ap <- sum(data_test$asthma_status == 1)

specificity = tp/ap
specificity

# Calculate sensitivity
tn <- sum(logit_bin_pred ==0 & data_test$asthma_status == 0)
an <- sum(data_test$asthma_status == 0)

sensitivity <- tn/an
sensitivity
```

Extracting and graphing largest coefficients in model. 

```{r}
mdl_coeff <- as.data.frame(coefficients(mdl))

mdl_coeff <- abs(mdl_coeff) %>%
  mutate(gene = rownames(.)) %>%
  clean_names() %>%
  filter(gene != "(Intercept)") %>%
  arrange(desc(coefficients_mdl)) %>%
  dplyr::slice(c(1:10))

theme_set(theme_bw())

ggplot(mdl_coeff, aes(x = gene, y = coefficients_mdl)) +
  geom_col() +
  scale_x_discrete(limits=mdl_coeff$gene) + 
  labs(title = "Logistic Regression: Top 10 Proteins for Separation by Asthma Status", x="Gene", y = "Coefficient")
```

