---
title: "PREHEAT Data Analysis Report"
author: "Elise Hickman, Hannah Matthews, Morgan Nalesnik"
date: "2022-10-21"
output: 
 html_document:
    toc: true
    toc_float: true
    toc_depth: 5
    code_folding: show
---

## PREHEAT Retreat Script Description

Authored by Elise Hickman, Hannah Matthews, and Morgan Nalesnik. Group 5 - Asthma Status.
Aim of script: determine whether there are significant protein expression differences between asthmatics and nonasthmatics.

## Two Group Comparisons

#### Setup and Data Import

```{r message = FALSE}
# Load packages
library(tidyverse)
library(ggplot2)
library(reshape2)
library(Hmisc)
library(superheat)
```

Load data.
```{r}
# Demographic Data
demo.data <- read.csv("../Data/Demographic_Data.csv")

# Proteomics Data
proteomics.data <- read.csv("../Data/Processed_Proteomics_Data.csv")

# Make sure everyone has two samples
table(demo.data$Subject_Number)
table(demo.data$PrevPost)
```
#### Data Exploration
```{r}
describe(demo.data)
```


Plotting boxplots of all genes per sample, same as what we did in pre-processing script. Data needs to be reformatted to tidy format.

```{r}
proto.long = proteomics.data %>% pivot_longer(cols=-c("Gene_Name"))
proto.long$PrevPost = gsub("_.*", "", proto.long$name)
ggplot(data=proto.long, aes(x=name, y=log10(value), col=PrevPost)) + geom_boxplot(outlier.shape = NA)
```
We can also use a heatmap to visualize all the data at once using superheat package.

```{r}
superheat(log(proteomics.data[,2:ncol(proteomics.data)]), pretty.order.rows = T, 
          bottom.label.text.angle = 90)
```

#### Dataset manipulation

General data manipulation/organization.
```{r}
# Combine data
proteomics.data = proteomics.data %>% filter(!is.na(Gene_Name)) %>%   # filter lets us filter by some condition, here Gene_Name that is NOT NA (aka not missing)

mutate(Gene_Name = gsub("-", "_", Gene_Name))  # mutate allows us to create a new variable or modify existing one, here I'm just changing - to _ in gene names

# First transpose the data so subjects are on rows
proto.tr.data = data.frame(t(proteomics.data)) # "t" transposes dataset (columns to rows, rows to columns)
colnames(proto.tr.data) = proto.tr.data[1,]
proto.tr.data = proto.tr.data[-1,]

# Create ID column
proto.tr.data$SampleID = rownames(proto.tr.data)


# Merge two datasets so everything is together
combined.data = merge(demo.data, proto.tr.data, by = "SampleID")
combined.data = data.frame(combined.data)

long.data.pre = combined.data %>% filter(PrevPost == "Pre") %>%     # filter to only "Pre" exposure
  dplyr::select(c("Subject_Number", proteomics.data$Gene_Name)) %>%        # select alows you to just use the specified columns
  pivot_longer(cols = -c(Subject_Number), names_to = "Gene", values_to = "Value_Pre") # pivot longer turns the data into long form
                                                                                       # cols says which columns to pivot longer -- i am telling it to use everything EXCEPT Subject_Number
                                                                                       # names_to  says where to store info from column names for data specified by "cols"
                                                                                       # values_to  says where to store the actual data

# Do the same thing for post exposure data

long.data.post = combined.data %>% filter(PrevPost == "Post") %>% 
  dplyr::select(c("Subject_Number", proteomics.data$Gene_Name)) %>%
  pivot_longer(cols = -c(Subject_Number), names_to = "Gene", values_to = "Value_Post")


# Combine pre and post by subject number

long.data = merge(long.data.pre, long.data.post, by = c("Gene", "Subject_Number"))
```

#### Asthma-Associated Proteins

More data cleaning.

```{r}
# For all these questions we are looking at PRE exposure, so let's just create a long dataset with only pre exposure across the various columns

long.data.pre = combined.data %>% filter(PrevPost == "Pre") %>% 
  dplyr::select(c("Subject_Number", proteomics.data$Gene_Name, "Responder_Status", "Sex", "Race", "Age", "BMI", "Asthma_Status")) %>%
  pivot_longer(cols = -c(Subject_Number, Responder_Status, Sex, Race, Age, BMI, Asthma_Status), names_to = "Gene", values_to = "Value_Pre") %>%
  arrange(Gene)
long.data.pre$Value_Pre = as.numeric(long.data$Value_Pre)


long.data.pre.loop = combined.data[which(combined.data$PrevPost=="Pre"),]

long.data.post = combined.data %>% filter(PrevPost == "Post") %>% 
  dplyr::select(c("Subject_Number", proteomics.data$Gene_Name, "Responder_Status", "Sex", "Race", "Age", "BMI", "Asthma_Status")) %>%
  pivot_longer(cols = -c(Subject_Number, Responder_Status, Sex, Race, Age, BMI, Asthma_Status), names_to = "Gene", values_to = "Value_Post") %>%
  arrange(Gene)
long.data.post$Value_Post = as.numeric(long.data$Value_Post)

long.data.post.loop = combined.data[which(combined.data$PrevPost=="Post"),]

long.data = merge(long.data.pre, long.data.post, by = c("Gene", "Subject_Number"))

```


# Pre-Exposure: What proteins differ by asthma status?

```{r}
stat.test.asthma.pre = long.data.pre %>%
  group_by(Gene) %>%
  summarise(pval = t.test(Value_Pre ~ Asthma_Status)$p.value,
            as_tibble(as.list(t.test(Value_Pre ~ Asthma_Status)$estimate)),
            statistic = t.test(Value_Pre ~ Asthma_Status)$statistic) %>%
  mutate(pval.BH = p.adjust(pval, method="BH"))

stat.test.asthma.pre.order = stat.test.asthma.pre[order(stat.test.asthma.pre$pval.BH),]

combined.data.pre <- filter(combined.data, PrevPost == 'Pre')
combined.data.pre$HLA_C <- as.numeric(combined.data.pre$HLA_C)

theme_set(theme_bw())
ggplot(combined.data.pre, aes(x = Asthma_Status, y = HLA_C)) +
  geom_boxplot() +
  geom_jitter(position=position_jitter(width=0.1, height=0.1)) +
  labs(title = "HLA-C Expression", x = "Asthma Status", y = "Relative Expression")
```

# Pre-Exposure: What proteins differ by asthma status?
```{r}
stat.test.asthma.post = long.data.post %>%
  group_by(Gene) %>%
  summarise(pval = t.test(Value_Post ~ Asthma_Status)$p.value,
            as_tibble(as.list(t.test(Value_Post ~ Asthma_Status)$estimate)),
            statistic = t.test(Value_Post ~ Asthma_Status)$statistic) %>%
  mutate(pval.BH = p.adjust(pval, method="BH"))

stat.test.asthma.post.order = stat.test.asthma.post[order(stat.test.asthma.post$pval.BH),]

combined.data.post <- filter(combined.data, PrevPost == 'Post')
combined.data.post$HEXA <- as.numeric(combined.data.post$HEXA)

theme_set(theme_bw())
ggplot(combined.data.post, aes(x = Asthma_Status, y = HEXA)) +
  geom_boxplot() +
  geom_jitter(position=position_jitter(width=0.1, height=0.1)) +
  labs(title = "HEXA Expression", x = "Asthma Status", y = "Relative Expression")
```

